Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model

### 2015-04-29 msc-publication.tex
Add citations for OpenFOAM & add2dMountain

\abstract{Terrain following coordinates are widely used in operational models but the cut cell method has been proposed as an alternative that can more accurately represent the orography and associated atmospheric dynamics.  Because the type of grid is usually chosen during model implementation, it is typically necessary to use different models in order to compare the accuracy of different grids.  On the contrary, we use a single C-grid finite volume model to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test of a stably stratified atmosphere at rest, and orographically induced gravity waves.  In addition, we formulate a new tracer advection test having a velocity field that is everywhere tangential to the terrain following coordinate surfaces.
The results of the advection tests demonstrate that tracer accuracy depends upon alignment of the flow with the grid.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature.  However, we find that the Lorenz computational mode is excited as a vertical zig--zag in potential temperature on the cut cell grid.
\TODO{this last sentence is weak: Although errors do vary between grids, overall, a high degree of accuracy is obtained for all tests on all grids.}}

@@

\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h \label{eqn:btf}
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the physical height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.  An example is shown in figure~\ref{fig:grid}a.

@@

The hybrid terrain following (HTF) coordinates of \citet{simmons-burridge1981} improve upon BTF coordinates by using a decay function that allows the influence of the terrain to be removed at a specified height, producing flat model layers aloft.  \TODO{I could mention that this is used operationally in UK MO (davies et al 2005) and ECMWF (good citation needed)}

@@

The OpenFOAM utility \texttt{snappyHexMesh} was used to create a grid that approximates the cut cell method.  First, a custom utility was used to move points beneath the surface up to the surface creating small cells near mountain peaks.  Second, the surface faces were taken from the BTF grid and \texttt{snappyHexMesh} was used to intersect the surface with the grid.  The tool removes cells whose centres are below the surface and displaces boundary vertices so that they lie on the BTF surface.  An example of the resulting grid is shown in figure~\ref{fig:grid}c.

@@

The flux form of the advection equation, \(\partial \phi / \partial t + \nabla \cdot \left( \bm{u} \phi \right) = 0\), is solved using an upwind-biased cubic advection scheme which is non-monotonic and not flux corrected.  The time derivative is solved using a three-stage, second order Runge-Kutta scheme:
\begin{align}
	\varphi^\star &= \varphi^{(n)} + \Delta t f(\phi^{(n)}) \\
	\varphi^{\star\star} &= \varphi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^\star) \right) \\
	\varphi^{(n+1)} &= \varphi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^{\star\star}) \right)
\end{align}
where \(f(\phi^{(n)}) = - \mathrm{div}(\bm{u} \varphi^{(n)})\) at time level \(n\).

@@

\begin{table*}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests using centred linear and cubic upwind-biased schemes.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{ r @{\hspace{0.5em}} l l l | @{\hspace{0.2em}} l l l | @{\hspace{0.5em}} l l l | @{\hspace{0.2em}} l l l}
\hline\hline
& \multicolumn{6}{c}{Horizontal advection} & \multicolumn{6}{c}{Terrain following advection} \\
& \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} \\
& \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max \\
\hline
Analytic  & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 \\
BTF 	  & \num{0.0318} & \num{-0.442} & \num{0.849} & \num{0.00767} & \num{-0.0433} & \num{0.928} & \num{0.00251} & \num{-0.0244} & \num{0.985} & \num{0.00154} & \num{-0.0105} & \num{0.983} \\
SLEVE 	  & \num{0.00233} & \num{-0.0253} & \num{0.945} & \num{0.00108} & \num{-0.0106} & \num{0.981} & \num{0.0298} & \num{-0.258} & \num{0.865} & \num{0.0122} & \num{-0.0273} & \num{0.864} \\
Cut cell  & \num{0.00224} & \num{-0.0253} & \num{0.985} & \num{0.000579} & \num{-0.00670} & \num{0.982} & \num{0.0318} & \num{-0.270} & \num{0.846} & \num{0.0136} & \num{-0.0297} & \num{0.850} \\
No terrain & \num{0.00224} & \num{-0.0253} & \num{0.985} & \num{0.000577} & \num{-0.00670} & \num{0.982} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & {---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
\hline
\end{tabular}
\end{table*}

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-gravityWaves/fig-gravityWaves.pdf}
%
	\caption{Vertical cross section of vertical velocity contours in the gravity waves test after 5 hours with mountain height $h_0 = \SI{250}{\meter}$ on (a) the BTF grid compared with (b) the mass-conserving semi-implicit semi-Lagrangian solution from \citet{melvin2010}.  Contours are every \SI{5e-2}{\meter\per\second} with solid lines denoting ascent and dashed lines descent.}
	\label{fig:gw-w}
\end{figure}

>>>>

\abstract{Terrain following coordinates are widely used in operational models but the cut cell method has been proposed as an alternative that can more accurately represent the orography and associated atmospheric dynamics.  Because the type of grid is usually chosen during model implementation, it is typically necessary to use different models in order to compare the accuracy of different grids.  On the contrary, we use a single C-grid finite volume model to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test of a stably stratified atmosphere at rest, and orographically induced gravity waves.  In addition, we formulate a new tracer advection test having a velocity field that is everywhere tangential to the terrain following coordinate surfaces.
The results of the advection tests demonstrate that tracer accuracy depends upon alignment of the flow with the grid.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature.  However, we find that the Lorenz computational mode is excited as a vertical zig--zag in potential temperature on the cut cell grid.}

@@

\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h \label{eqn:btf}
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the physical height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  This formulation results in $z^\star/H$ ranging from 0 to 1.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.  An example is shown in figure~\ref{fig:grid}a.

@@

The hybrid terrain following (HTF) coordinates of \citet{simmons-burridge1981} improve upon BTF coordinates by using a decay function that allows the influence of the terrain to be removed at a specified height, producing flat model layers aloft.  \TODO{include equation?}

@@

The OpenFOAM utility \texttt{snappyHexMesh} was used to create a grid that approximates the cut cell method.  First, a custom utility was used to move points beneath the surface up to the surface creating small cells near mountain peaks.  The utility is available at \url{https://github.com/hertzsprung/AtmosFOAM/tree/2386ffb/applications/utilities/mesh/add2dMountain}.  Second, the surface faces were taken from the BTF grid and \texttt{snappyHexMesh} was used to intersect the surface with the grid.  The tool removes cells whose centres are below the surface and displaces boundary vertices so that they are `snapped' to the BTF surface \citep{openfoam2015}.  An example of the resulting grid is shown in figure~\ref{fig:grid}c.

@@

The flux form of the advection equation, \(\partial \phi / \partial t + \nabla \cdot \left( \bm{u} \phi \right) = 0\), is solved using the upwind-biased cubic advection scheme described by \cite{weller-shahrokhi2014} which is non-monotonic and not flux corrected.  The time derivative is solved using a three-stage, second order Runge-Kutta scheme:
\begin{align}
	\varphi^\star &= \varphi^{(n)} + \Delta t f(\phi^{(n)}) \\
	\varphi^{\star\star} &= \varphi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^\star) \right) \\
	\varphi^{(n+1)} &= \varphi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^{\star\star}) \right)
\end{align}
where \(f(\phi^{(n)}) = - \mathrm{div}(\bm{u} \varphi^{(n)})\) at time level \(n\).

@@

\begin{table*}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests using centred linear and cubic upwind-biased schemes.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{ r @{\hspace{0.5em}} l l l | l l l | l l l | l l l}
\hline\hline
& \multicolumn{6}{c}{Horizontal advection} & \multicolumn{6}{c}{Terrain following advection} \\
& \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{3}{c}{Centred linear} & \multicolumn{3}{c}{Cubic upwind-biased} \\
& \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max & \(\ell^2\) error & min & max \\
\hline
Analytic  & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 & 0 & 0 & 1 \\
BTF 	  & \num{0.0318} & \num{-0.442} & \num{0.849} & \num{0.00767} & \num{-0.0433} & \num{0.928} & \num{0.00251} & \num{-0.0244} & \num{0.985} & \num{0.00154} & \num{-0.0105} & \num{0.983} \\
SLEVE 	  & \num{0.00233} & \num{-0.0253} & \num{0.945} & \num{0.00108} & \num{-0.0106} & \num{0.981} & \num{0.0298} & \num{-0.258} & \num{0.865} & \num{0.0122} & \num{-0.0273} & \num{0.864} \\
Cut cell  & \num{0.00224} & \num{-0.0253} & \num{0.985} & \num{0.000579} & \num{-0.00670} & \num{0.982} & \num{0.0318} & \num{-0.270} & \num{0.846} & \num{0.0136} & \num{-0.0297} & \num{0.850} \\
No terrain & \num{0.00224} & \num{-0.0253} & \num{0.985} & \num{0.000577} & \num{-0.00670} & \num{0.982} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & {---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
\hline
\end{tabular}
\end{table*}

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-gravityWaves/fig-gravityWaves.pdf}
%
	\caption{Vertical cross section of vertical velocity contours in the gravity waves test after 5 hours with mountain height $h_0 = \SI{250}{\meter}$ on (a) the BTF grid compared with (b) the mass-conserving semi-implicit semi-Lagrangian solution from \citet{melvin2010}.  Contours are every \SI{5e-2}{\meter\per\second} with solid lines denoting ascent and dashed lines descent.  Small oscillations in contour lines are due to contouring on an irregular grid.}
	\label{fig:gw-w}
\end{figure}
